---
title: "SD Single Cell Genotyping Data Reduction"
output:
  html_document: default
  pdf_document: default
---
*****Click on the gear above and choose:  "Chunk Output in Console" before proceeding*****

## Initialize libraries and Directory setup
```{r Load yer libraries, echo=FALSE, message=FALSE}
require(plyr); require(dplyr); require(tidyr); require(reshape2); require(ggplot2);
require(ggExtra); require(ggrepel); require(gridExtra);  require(RColorBrewer); require(stringr);
options(stringsAsFactors = FALSE)
```


Sets up the directory structure for where to look for data, and where to save results.  
This assumes that the data structure is what is in the github repo and what is produced by the ImageJ macros. 
```{r Directories, echo=FALSE}
output <- c("./OutputData/")
dir.create(output, showWarnings = FALSE)
metadataDir <- c("./ProcessedData/Metadata")
fluorInput <- c("./ProcessedData/TyphoonCSVs")
cellInput <- c("./ProcessedData/OlympusCSVs")
writeLines(capture.output(sessionInfo()), 
           paste(output,"/RCrunching-sessionInfo.txt", sep = ""))
```

## Run Metadata 
```{r Metadata}
arrayList <- c("A1", "A2", "A3")
ampFluor <- "FAM"; wtFluor <- "Cy5"; mutFluor <- "HEX"
fluorList <- c(ampFluor, wtFluor, mutFluor)
```

# Read in and Melt Raw Data
Read in the individual csv's for each fluorphore, via searching the input
directory for filenames containing the names of the fluors used in this 
experiment.  This generates a melted data frame containing the concatenated
data sets from each fluorophore.  
```{r melt all data}
fileNames <- list.files(path = fluorInput, pattern = "*.csv", full.names = TRUE)
dataList <- lapply(fluorList, function(x) 
    {read.csv(file = fileNames[grep(x, fileNames)], stringsAsFactors = FALSE)})
names(dataList) <- fluorList
dataList <- lapply(dataList, function(x) {colnames(x)[1] <- "Well";x}) 
allFluors <- melt(dataList, id=colnames(dataList[[1]]))
colnames(allFluors)[length(colnames(allFluors))] <- "Fluor"
names(allFluors)[names(allFluors) == "X.Area"] <- "Area"

rm(dataList)
```

## Calculate AID
Note: Area is divided by Max % Fill per ROI, which has already been calculated 
for this master/array design as 9 pixels by 17 pixels, and ROI grid is 400 pixels, 
so the max % Fill per ROI is 38.25%
```{r define arrays and calculate AID}
wellTotal <- 1024  # Total number of wells in a single array
allFluors$Array[allFluors$Well <= wellTotal] <- "A1"
allFluors$Array[allFluors$Well >= wellTotal+1 & allFluors$Well <= 2*wellTotal] <- "A2"
allFluors$Array[allFluors$Well > 2*wellTotal] <- "A3"

allFluors$AID <- allFluors$IntDen/(allFluors$Area/38.25)
allFluors$Fluor <- factor(allFluors$Fluor, levels = fluorList)
allFluors$Array <- factor(allFluors$Array)
allFluors <- allFluors %>% mutate(Template = "OCI-AML3")
```

## Data formatting to generate a "spread" format for further analysis.  
Uses the area calculated in the ampFluor channel as the true Area from here on out.
```{r}
procData <- allFluors[c("Well", "Fluor", "Array", "AID")]
ampArea <- allFluors %>% filter(Fluor == ampFluor) %>% select(Well, Area, Array, Template)

wideForm <- procData %>% group_by(Array) %>% spread(Fluor, AID) 
colnames(wideForm)[which(grepl(ampFluor, colnames(wideForm)))] <- "ampFluor"
colnames(wideForm)[which(grepl(wtFluor, colnames(wideForm)))] <- "wtFluor"
colnames(wideForm)[which(grepl(mutFluor, colnames(wideForm)))] <- "mutFluor"

primeTime <- merge(ampArea, wideForm)
primeTime <- primeTime[order(primeTime$Well),] 
primeTime$ArrayRow <- rep(1:16, each = 64)
primeTime$ArrayCol <- seq(1:64)

rm(procData, ampArea, wideForm)
```

## Fill Thresholding
Plot data as a function of area in order to visualize the area threshold for 
identifying which wells "filled" and which did not prior to downstream gating.

Visualize the suggested threshold as well as the chosen threshold to confirm.
```{r}
tArea <- 21
ggplot(data = primeTime %>% filter(Area > 5), aes(x=Area)) +
    geom_density() + facet_grid(. ~ Array) + 
    geom_vline(xintercept = tArea, color = "springgreen4") +
    theme_bw()
```

Apply one area threshold to the data frame for all arrays and generate a new 
"ruleThemAll". 
```{r}
ruleThemAll <- primeTime %>% group_by(Array) %>% mutate(Filled = Area > tArea)
```

## HEX Correction and Thresholding
Set the constant used to correct for FAM bleed-through in the HEX channel.  This value was calcuated based on the average of several runs of mutant plasmids.
```{r HEX Correction}
avgHEXSlope <- 0.46
#snowCone <- mutate(ruleThemAll, runID = "OCI-AML3 Cells")

snowCone <- mutate(ruleThemAll, mutFluorAdj = mutFluor - avgHEXSlope*ampFluor)
#snowCone$mutFluorAdj <- NA
```

Required for thresholding: 
Calculate max of density curve
```{r calculate maximums}
#set approximate range of negative wells to get more accurate denisity curve
AmpMax    <- 1100000
MutMax    <- 2000000
WtMax     <- 1500000
MutAdjMax <- 1000000
WtAdjMax  <- 1500000

#rearrange columns to make gather work
snowCone <- select(snowCone, Well, Array, Area, Template, ampFluor, wtFluor, mutFluor, mutFluorAdj, ArrayRow, ArrayCol, Filled)

#function to find the max density
Peak <- function(X) {
      Y = density(X, na.rm = TRUE); 
      Z = data.frame(Intensity = Y$x, Density = Y$y);
    return(Z$Intensity[which.max(Z$Density)])
}

#make a table with max AID's for each Fluor
snowPeaks <- snowCone %>% filter(ampFluor < AmpMax, mutFluor < MutMax, wtFluor < WtMax) %>% gather(Fluor, AID, ampFluor:mutFluor) %>% 
          filter(Filled == TRUE) %>% 
          group_by(Array, Fluor) %>% 
          do(data.frame(maxAID = Peak(.$AID)))

mutAdjPeaks <- snowCone %>% gather(Fluor, AID, ampFluor:mutFluorAdj) %>% 
          filter(Filled == TRUE, Fluor == "mutFluorAdj", AID < MutAdjMax) %>% 
          group_by(Array, Fluor) %>% 
          do(data.frame(maxAID = Peak(.$AID)))

snowPeaksAdj <- bind_rows(snowPeaks, mutAdjPeaks)
```

Calculate thresholds using previously determined maxSD of negative controls
```{r calculate thresholds}
xsigamp <- 3
xsigmut <- 6
xsigwt <- 5

PSwapampSD    <- 41753*xsigamp
PSwapwtSD     <- 29937*xsigwt
PSwapmutAdjSD <- 52998*xsigmut

#Add thresholds to snowPeaksAdj
snowDrift <- snowPeaksAdj %>% mutate(Threshold = if_else( Fluor == "ampFluor", maxAID + PSwapampSD,
                                                  if_else( Fluor == "wtFluor", maxAID + PSwapwtSD,
                                                  if_else( Fluor == "mutFluorAdj", maxAID + PSwapmutAdjSD,
                                                          0))))

#Add this threshold to the df with all the AID's
snowDrift1 <- snowDrift %>% select(Array, Fluor, Threshold) %>% spread(Fluor, Threshold) %>% select(Array, ampThresh = ampFluor, mutThresh = mutFluor, mutAdjThresh = mutFluorAdj, wtThresh = wtFluor)

snowCone1 <- left_join(snowCone, snowDrift1, by = c("Array"))

```

Visualize amplification thresholds to spot errors
```{r view amp thresh}
ampThreshPlot <- ggplot(snowCone1 %>% filter(Filled == TRUE), aes(x=ampFluor)) + 
                            geom_histogram(aes(x = ampFluor, y = ..density..), bins = 20) +
                            geom_point(stat = "density", alpha = 0.6, stroke = 0) + 
                            geom_vline(aes(xintercept = ampThresh), color = "orchid4") +
                            xlim(500000, 1500000) +
                            xlab("AID") + ylab("Density") + 
                            facet_grid(.~ Array) +
                            ggtitle("amp Thresh") +
                            theme_bw() +
                            theme(strip.text.y = element_blank(), axis.text.x = element_text(angle=90, hjust=1))
ampThreshPlot
```

View scatter plot of mutant fluorophore vs wild-type fluorophore intensity
```{r Preliminary View of wt and mut Thresholds}
mutAdj_vs_wt_Scatter <- ggplot(snowCone1 %>% filter(Filled == TRUE), 
            aes(x=wtFluor, y=mutFluorAdj)) + 
            geom_point(alpha = 0.6, stroke = 0, color = "orangered2") + 
            geom_vline(aes(xintercept = wtThresh), color = "orchid4") +
            geom_hline(aes(yintercept = mutAdjThresh), color = "orchid4") +
            xlim(0,max(snowCone1$wtFluor, na.rm = TRUE)) +
            ylim(0,max(snowCone1$mutFluorAdj, na.rm = TRUE)) +
            labs(x = "wtFluor", y = "mutFluorAdj") +
            facet_grid(. ~ Array) +
            ggtitle("Well Intensities, with HEX Correction") +
            theme_bw() +
            theme(strip.text.y = element_blank(), axis.text.x = element_text(angle=90, hjust=1))
mutAdj_vs_wt_Scatter
```

Save a .jpeg of the mut vs. wt scatter plots
```{r save plots, messages=FALSE}
ggsave(mutAdj_vs_wt_Scatter, filename = paste0(output,"mutAdj_vs_wt_Scatter.jpeg"))
```

Determine amplification positivity for each well
```{r Apply amp Threshold}
snowCone2 <- mutate(snowCone1, AmpPos = FALSE)

snowCone2 <- snowCone2 %>% mutate(AmpPos = ampFluor >= ampThresh)

snowCone2$AmpPos <- as.factor(snowCone2$AmpPos)
```

Determine Zygosities in the individual arrays
```{r Apply wt and mut Thresholds}
snowCone3 <- mutate(snowCone2, Zygosity = "UNCALLED")

snowCone3$wtFluor[is.na(snowCone3$wtFluor)] <- 0
snowCone3$mutFluor[is.na(snowCone3$mutFluor)] <- 0
snowCone3$mutFluorAdj[is.na(snowCone3$mutFluorAdj)] <- 0
snowCone3$ampFluor[is.na(snowCone3$ampFluor)] <- 0


snowCone3 <- snowCone3 %>% mutate(Zygosity = replace(Zygosity, Filled == TRUE & AmpPos == TRUE  & 
                     wtFluor >= wtThresh & mutFluorAdj < mutAdjThresh, "WT"))

snowCone3 <- snowCone3 %>% mutate(Zygosity = replace(Zygosity, Filled == TRUE & AmpPos == TRUE  &
                     wtFluor < wtThresh & mutFluorAdj >= mutAdjThresh, "MUT"))

snowCone3 <- snowCone3 %>% mutate(Zygosity = replace(Zygosity, Filled == TRUE & AmpPos == TRUE  &
                     wtFluor >= wtThresh & mutFluorAdj >= mutAdjThresh, "HET"))

snowCone3$Zygosity <- factor(snowCone3$Zygosity, levels = c("WT", "HET", "MUT","UNCALLED"))

snowCone3 <- mutate(snowCone3, FluorQC = factor("ThisIsFine", levels = c("wtFluor", "mutFluor", "Both", "AmpOnly", "Negative", "ThisIsFine")))


snowCone3 <- snowCone3 %>% mutate(FluorQC = replace(FluorQC, Filled == TRUE & AmpPos == FALSE  &
                     wtFluor >= wtThresh & mutFluorAdj < mutAdjThresh, "wtFluor"))

snowCone3 <- snowCone3 %>% mutate(FluorQC = replace(FluorQC, Filled == TRUE & AmpPos == FALSE  & 
                     wtFluor < wtThresh & mutFluorAdj >= mutAdjThresh, "mutFluor"))

snowCone3 <- snowCone3 %>% mutate(FluorQC = replace(FluorQC, Filled == TRUE & AmpPos == FALSE & 
                     wtFluor >= wtThresh & mutFluorAdj >= mutAdjThresh, "Both"))

snowCone3 <- snowCone3 %>% mutate(FluorQC = replace(FluorQC, Filled == TRUE & AmpPos == TRUE & 
                     wtFluor < wtThresh & mutFluorAdj < mutAdjThresh, "AmpOnly"))

snowCone3 <- snowCone3 %>% mutate(FluorQC = replace(FluorQC, Filled == TRUE & AmpPos == FALSE & 
                     wtFluor < wtThresh & mutFluorAdj < mutAdjThresh, "Negative"))
```

## Plot Zygosity Results
Plot mut vs. wt with Zygosity
```{r View Results of wt and mut Thresholding}
FinalZygosity_PSwap <- ggplot(snowCone3 %>% filter(Filled ==TRUE, AmpPos == TRUE | FluorQC == "Negative"), 
            aes(x=wtFluor, y=mutFluorAdj)) + 
            geom_point(aes(color = Zygosity),alpha = 0.4, stroke = 0) + 
            geom_vline(aes(xintercept = wtThresh), color = "springgreen4") +
            geom_hline(aes(yintercept = mutAdjThresh), color = "springgreen4") +
        scale_color_manual(values = c("dodgerblue4","magenta4", "red3", "black"), limits = levels(snowCone3$Zygosity)) +
            xlim(0,max(snowCone3$wtFluor, na.rm = TRUE)) +
            ylim(-200000,max(snowCone3$mutFluorAdj, na.rm = TRUE)) +
            labs(x = "wtFluor", y = "mutFluorAdj") +
            facet_grid(. ~ Array) +
            ggtitle("PSwap Zygosity") +
            theme_bw() +
            theme(axis.text.x = element_text(angle=90, hjust=1), strip.text.y = element_text(size = 7, angle = 0))
FinalZygosity_PSwap
```

Save a .jpeg of the mut vs. wt Zygosity and FluorQC scatter plots
```{r save scatter plots, messages=FALSE}
ggsave(FinalZygosity_PSwap, filename = paste0(output, "FinalZygosity_PSwap.jpeg"))
```

## Cell Calculations

Read In Cell Data
```{r Read in Cell Data}
filenames <- list.files(path = cellInput, pattern = ".csv", full.names = T)
images <- as.numeric(str_sub(gsub(".csv", "", filenames), -2,-1))
cellData <- lapply(filenames, function(x) read.csv(x, header = T))
names(cellData) <- images
cellMelt <- melt(cellData, id = colnames(cellData[[1]]))
colnames(cellMelt) <- c("imageWell", "cellCount", "imageId")
cellMelt$ROIorder <- seq(1:3072)

lookup <- data.frame(ROIorder = 1:3072)
lookup$ArrayCol <- rep(c(rep(1:11, len = 88), rep(12:22, len = 88), 
                         rep(23:33, len = 88), rep(34:44, len = 88), 
                         rep(45:55, len = 88), rep(56:64, len = 144),
                         rep(45:55, len = 88), rep(34:44, len = 88), 
                         rep(23:33, len = 88), rep(12:22, len = 88), 
                         rep(1:11, len = 88)), 3)
lookup$ArrayRow <- rep(c(rep(1:8, each = 11, len = 440), 
                         rep(1:8, each = 9),rep(9:16, each = 9), 
                         rep(9:16, each = 11, len = 440)),3)
lookup$estImageId <- c(rep(1:5, each = 88), rep(6:7, each = 72), 
                       rep(8:17, each = 88), rep(18:19, each = 72), 
                       rep(20:29, each = 88), rep(30:31, each = 72), 
                       rep(32:36, each = 88))
lookup$Array <- rep(c("A1", "A2", "A3"), each = 1024)
lookup <- arrange(lookup, Array, ArrayRow, ArrayCol)
lookup$Well <- seq(1:3072)
fullMergeData <- join(lookup, cellMelt, by = "ROIorder", type = "left")
cellsReadyToRoll <- select(fullMergeData, Well, cellCount)
snowCone4 <- left_join(snowCone3, cellsReadyToRoll, by = c("Well"))
```

View the zygosities of the single cells per each array
```{r Plot zygosities of cells}
set1 <- brewer.pal(n = 9, "Set1")
set3 <- brewer.pal(n = 12, "Set3")
zygcolors <- c(set3[5], set3[10], set3[4], "black", "white")

zygBarPlot <- ggplot(data = snowCone4 %>% filter(Zygosity != "UNCALLED" & cellCount == 1),
                     aes(x = Zygosity, fill = Zygosity, group = Array), color = "black") +
                      geom_bar(aes(y = ..prop.., fill = factor(..x..))) + 
                      ylim(0, 1) +
                      scale_fill_manual(values = zygcolors) + theme_bw() +
                      geom_text(aes(label = ..count.., y = ..prop..), stat = "count", vjust = -0.5) +
                      labs(x = "Zygosity", y = "Frequency") + 
                      scale_x_discrete(limits=c("WT", "HET", "MUT")) +
                      theme(legend.position = "none") +
                      facet_grid(. ~ Array) +
                      theme(strip.text.y = element_text(size = 7, angle = 0))
zygBarPlot
```

Save the zygosity plot
```{r Save zygBarPlot and dropoutTable, messages=FALSE}
ggsave(paste0(output, "zygBarPlot-singleCells.jpg"),
       zygBarPlot,
       width = 6, height = 6)
```

## Failure rate calculations
```{r Failure Rates}
failureData <- snowCone4 %>% group_by( Array, Template) %>%
    summarise(FalsePositives = n_distinct(which(Filled == TRUE & AmpPos == TRUE & cellCount == 0)),
              FalseNegatives = n_distinct(which(Filled == TRUE & AmpPos == FALSE & cellCount == 1)),
              TruePositives = n_distinct(which(Filled == TRUE & AmpPos == TRUE & cellCount == 1)),
              TrueNegatives = n_distinct(which(Filled == TRUE & AmpPos == FALSE & cellCount == 0)),
              DoubletPlus = n_distinct(which(Filled == TRUE & cellCount > 1)),
              EmptyWells = n_distinct(which(Filled == FALSE)),
              FDR = FalsePositives/(FalsePositives + TruePositives),
              FPR = FalsePositives/(FalsePositives + TrueNegatives),
              FNR = FalseNegatives/(FalseNegatives + TruePositives)
              )
```

View stacked bar plot of failure data
```{r Plot Failure Rates}
failColors <- brewer.pal(6, "Accent")
failurePlotCounts <- melt(failureData %>% select(-c(FDR, FPR, FNR)), variable.name = "Mode", value.name = "WellCount")
failurePlot <- ggplot(data = failurePlotCounts) + 
  geom_bar(aes(x = Array, y = WellCount, fill = Mode), stat = "identity") + 
  scale_fill_manual(values = failColors) +
  scale_y_continuous(expand = c(0, 10)) +
  theme_bw()
failurePlot
```

save plot
```{r Save failure Plot, messages=FALSE}
ggsave(paste0(output, "failurePlot.jpg"),
       failurePlot,
       width = 6, height = 6)
```

save csv
```{r Save stats data, messages=FALSE}
write.csv(failureData, file = paste0(output,"failureRateResults.csv"), row.names = F)
```

## Poisson Distribution calculations
```{r Poisson Calculations}
experimentalLambda <- snowCone4 %>% group_by(Array,Template) %>%
    summarise(TotalCounts =   n_distinct(which(Filled == TRUE)),
              ZeroCells =     n_distinct(which(Filled == TRUE & cellCount == 0)),
              OneCell =       n_distinct(which(Filled == TRUE & cellCount == 1)),
              DoubletPlus =   n_distinct(which(Filled == TRUE & cellCount > 1)),
              ExpLambda =    -log(ZeroCells/TotalCounts),
              PredSingles =  TotalCounts*ExpLambda*exp(-ExpLambda)/1,
              PredDoubPlus = TotalCounts - ZeroCells - PredSingles
              )
```

View stacked barplot of cell counts
```{r View how close to Poisson it is}
cellNumPlot <- melt(experimentalLambda %>% select(Array, ZeroCells, OneCell, DoubletPlus), variable.name = "Mode", value.name = "CellCount")
CellCountPlot <- ggplot(data = cellNumPlot) + 
  geom_bar(aes(x = Array, y = CellCount, fill = Mode), stat = "identity") +
  theme_bw()
CellCountPlot
```

Compare estimated and experimental zero/single/doublet+ counts, per array
```{r Cell Poisson}
PoissonTest <- experimentalLambda %>% gather("Occupancy", "Counts", ZeroCells:DoubletPlus) %>% select(Array:ExpLambda, Occupancy, Counts) %>% mutate(Origin = "Actual")
PoissonTest2 <- experimentalLambda %>% mutate(PredZero = ZeroCells) %>% select(Array:TotalCounts, ExpLambda, "OneCell"= PredSingles, "DoubletPlus" = PredDoubPlus, "ZeroCells" = PredZero) %>% gather("Occupancy", "Counts", OneCell:ZeroCells) %>% mutate(Origin = "Predicted")
PoissonFull <- PoissonTest %>% bind_rows(PoissonTest2)                                                                                                  

PoissonFit <- ggplot(data = PoissonFull, aes(x = Occupancy, y = Counts, fill = Origin)) + 
  geom_bar(stat = "identity", position = position_dodge()) + 
  scale_x_discrete(limits=c("ZeroCells", "OneCell", "DoubletPlus")) +
  facet_grid(.~ Array) +
  ggtitle("Actual and Poisson-Predicted Cell Counts") +
  theme_bw() +
  theme(axis.text.x = element_text(angle=90, hjust=1), strip.text.y = element_text(size = 7, angle = 0))
PoissonFit
```

Compare estimated and experimental zero/single/doublet+ counts, all data added together
```{r Poisson Summary}
PoissonSummary <- PoissonFull %>% group_by(Origin, Occupancy) %>% summarise(Counts = sum(Counts))

PoissonSummaryPlot <- ggplot(data = PoissonSummary, aes(x = Occupancy, y = Counts, fill = Origin)) + 
  geom_bar(stat = "identity", position = position_dodge()) + 
  scale_x_discrete(limits=c("ZeroCells", "OneCell", "DoubletPlus")) +
  ggtitle("Combined Actual and Poisson-Predicted Cell Counts") +
  theme_bw() +
  theme(axis.text.x = element_text(angle=90, hjust=1))
PoissonSummaryPlot
```

Save a .jpeg of the Poisson plots
```{r save Poisson plots, messages=FALSE}
ggsave(PoissonFit, filename = paste0(output,"PoissonFit.jpeg"))
ggsave(PoissonSummaryPlot, filename = paste0(output,"PoissonSummaryPlot.jpeg"))
ggsave(CellCountPlot, filename = paste0(output,"CellCountPlot.jpeg"))
```

Save csv's of cell count distribution data
```{r Save Poisson data, messages=FALSE}
write.csv(experimentalLambda, file = paste0(output,"poissionEstimateResults.csv"), row.names = F)
write.csv(PoissonFull, file = paste0(output,"PoissonFull.csv"), row.names = F)
write.csv(PoissonSummary, file = paste0(output,"PoissonSummary.csv"), row.names = F)
```

Save the final data frame
```{r write out cellsnowCone, messages=FALSE}
write.csv(snowCone4, file = paste0(output,"cellsnowCone.csv"), row.names = F)
```


